Network based multifactorial modelling microRNA-target interaction.

### 2019-09-17 article.tex
news to article
last meeting
article

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Network based multifactorial modelling microRNA-target interaction.},
            pdfauthor={Selcen Ari; Alper Yilmaz},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Network based multifactorial modelling microRNA-target interaction.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Selcen Ari\footnote{Department of Bioengineering, Yildiz Technical
  University, Istanbul, Turkey} \\ Alper Yilmaz\footnote{Department of Bioengineering, Yildiz Technical
  University, Istanbul, Turkey}}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{10 Eylül 2019}

@@

Competing endogenous RNA (ceRNA) regulations and crosstalk between
various types of non-coding RNA in human is remarkable in means of miRNA
regulation. Many studies have pointed out that an alteration in
miRNA:target interaction can result in unexpected changes due to
indirect and complex interactions. In this paper, we defined a new
network-based model that handles miRNA:ceRNA interactions with
expression values. Our model is able to handle miRNA interaction factors
such as seed type, binding energy, if provided. Our approach is able to
reveal that a perturbation in an element of network affects whole
competing elements differently and cooperative efficiencies of miRNAs on
common targets could be calculated. Our findings emphasized importance
of miRNA:target ratios being crucial, as reported by previous studies.
We have showed that the competing elements which have the same or close
expression values may not be affected equally from the perturbation
because of repression functionality depended on interaction factors of
miRNA target pairs (answer: Hocam etkileşim faktörlerinden dolayı
ekspresyonları yakın olsa bile baskılanma aktitesinin aynı
sergilenmeyeceğini ifade etmeye çalıştım. Bu şekilde anlaşılıyor mu?).
We applied the model to real sample consisting of breast cancer gene and
miRNA expression dataset and experimental miRNA:target interaction
dataset all generated via high throughput sequencing methods. A gene
over-expressed in tumor tissue, namely \emph{ABCC1}, is used as
perturbing element. We have observed that change in expression level of
single gene in miRNA:target network is sufficient to perturb regulations
in whole network, due to unforeseen and unpredicted regulation which are
only visible when considered in network context. Therefore, this model
helps unveiling the crosstalk between elements in miRNA:target network
where abundance of target and sponge effect are taken into account. The
model is scalable and can be plugged in with emerging miRNA effectors
such as circRNAs. The model is available as R package
\href{https://github.com/selcenari/ceRNAnetsim}{ceRNAnetsim}.

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key
regulator of gene expression through various post-transcriptional
mechanisms. Although the mechanisms by which miRNA represses are not
fully understood, miRNAs predominantly repress their targets. Repressive
activities of miRNAs vary depending on many factors that are significant
to miRNA:target interactions. These factors include miRNA:target binding
energy, binding location in target sequence, base pairing types between
miRNA and target, abundance of miRNAs and targets (Grimson et al.
\protect\hyperlink{ref-grimson_microrna_2007}{2007}). Binding energies
of miRNA:target complexes vary based on nucleotide context and determine
folding stability of complex (Cao and Chen
\protect\hyperlink{ref-cao_predicting_2012}{2012}). It has been
demonstrated that the binding energy between miRNA and target indicates
stability or affinity of complex (Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}) and does not directly
determine repressive activity of miRNA (Cao and Chen
\protect\hyperlink{ref-cao_predicting_2012}{2012}). Early studies have
argued that 2-8 nt sequence, seed, located in miRNA 5'end bind to
specific sequence located in 3'UTR of its target (Bartel
\protect\hyperlink{ref-bartel_micrornas_2004}{2004}; Lewis, Burge, and
Bartel \protect\hyperlink{ref-lewis_conserved_2005}{2005}). In recent
studies, it has been shown that miRNAs can interact with targets via
sequences located in regions such as 5'UTR or CDS (Hausser et al.
\protect\hyperlink{ref-hausser_analysis_2013}{2013}; Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}; Moore et al.
\protect\hyperlink{ref-moore_mirnatarget_2015}{2015}). These studies
also showed that binding location could indicate functionality of
miRNA:target interaction or be effective on abundance of targets(TOFIX
abundance etkisi derken, bazen azaltabilir, bazen arttırabilir,
manasında mı? answer: evet hocam anlaşılmıyorsa değiştirebilirim). It
has been shown that miRNAs exhibit repressive activity via, 6-8 nt long
sequence that is perfectly complementary with targets, seed region at
the 5' end of miRNAs (Bartel
\protect\hyperlink{ref-bartel_micrornas:_2009}{2009}; Grimson et al.
\protect\hyperlink{ref-grimson_microrna_2007}{2007}). On the other hand,
some researchers have reported that seed sequence of miRNA can have
mismatches or bulged/wobble nucleotides (Chi, Hannon, and Darnell
\protect\hyperlink{ref-chi2012alternative}{2012}) and may locate in
region other than 5'end of miRNAs (Hafner et al.
\protect\hyperlink{ref-hafner_transcriptome-wide_2010}{2010}; Helwak et
al. \protect\hyperlink{ref-helwak_mapping_2013}{2013}). On top of all
these factors, abundance of miRNAs and targets and miRNA:target ratio in
cells predominantly affect efficiency of miRNA:target interaction (Arvey
et al. \protect\hyperlink{ref-arvey_target_2010}{2010}; Bosson, Zamudio,
and Sharp \protect\hyperlink{ref-bosson_endogenous_2014}{2014}; Denzler
et al. \protect\hyperlink{ref-denzler_assessing_2014}{2014}).

@@

Initially, the network is assumed in steady-state (\autoref{fig1}a)
condition and needs least one trigger for initiating calculations. The
trigger can be a change in expression level of one or more genes
(\autoref{fig1}b). After a trigger, the network undergoes iterative
cycle of calculations at each of which distribution of miRNA in local
neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA
distribution, expression level of each node (i.e.~ceRNA) is updated.
This update results in change of expression value of common target (G4)
in system. In this case, the common target acts as trigger for the other
group (the targets interacted with M2 miRNA) in the network. Due to
common targeted elements, the change in one neighborhood spreads to
other neighborhoods (\autoref{fig1}d), consequently have potential to
effect whole network due to ``ripple effect''.

\begin{figure}
\hypertarget{fig1}{%
\centering
\includegraphics{Fig1.jpg}
\caption{Schematic presentation of mechanism of network based model. a)
In steady state, miRNAs (M: gray triangle) repress targets (G: circle
coconut) according to proportion of their targets' expression. b) Two
fold increase in transcript level of Gene2 (G2) acts as a trigger (shown
with red in figure). c) Distribution of miRNA1 (M1) changes. d) The
change at expression (shown with light red because minor changes are
occured on their expression values.) of common target effects changes of
proportional distribution of miRNA2 (M2). Expression values are rounded
to integers for simplicity}\label{fig1}
}
\end{figure}

@@

The repression efficiency of a miRNA on the individual target
(\(Eff_{gi}\)) is calculated according to equation \eqref{eq:1}; where
miRNA expression (\(C_m\)) in local neighborhood is distributed among
targets using individual gene expression levels (\(C_{gi}\)) (\#FIX
isn't \(C_g\) gene expressions in the group is equal to total of
\(C_{gi}\) in group? answer: its my false. now its corect.). For the
genes targeted by multiple miRNAs, cooperative activity of miRNAs on a
target gene, \(R\), is calculated by summing repression activity of each
miRNA (Equation \eqref{eq:2}).

@@

Interactions between miRNAs and their targets can be affected from
various factors. So, our model integrates multiple factors when
calculating overall miRNA activity. We classified factors into two
categories. Factors effecting binding determine interaction between
miRNA and target and they alter amount of miRNA sequestered to target.
Degradation efficiency factors assign amount of degradated target amount
in miRNA:target pairs.(\#FIX fix this sentence: answer:). In other
words, binding factors exert their influence before or during binding,
efficiency factors exert their influence after binding. In the
literature, binding free energy (Cao and Chen
\protect\hyperlink{ref-cao_predicting_2012}{2012} ; Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}) and seed type (Werfel
et al. \protect\hyperlink{ref-werfel_preferential_2017}{2017}) in
miRNA:target interactions are described as factors effecting binding
affinity. Efficiency factors determine how many of miRNA:target
complexes will result in inhibition and binding region on target
drastically effect miRNA degradation efficiency (Hausser et al.
\protect\hyperlink{ref-hausser_analysis_2013}{2013}; Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}). Both binding and
efficiency factors are normalized to their maximum values and scaled to
{[}0,1{]} interval. The normalized values of factors take into account
to determine binding activity and miRNA efficiency on targets
(\autoref{fig2}). Binding affinities (activity, \(Eff\)) of miRNAs on
each individual gene are calculated as shown in equation \eqref{eq:3};
\(C_m\), miRNA expression in the group; \(C_{gi}\), individual gene
expression; gi, individual gene (\autoref{fig2}c). \begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

\begin{figure}
\hypertarget{fig2}{%
\centering
\includegraphics{fig2.jpg}
\caption{Calculations to determine of miRNA binding and repression
efficiency. \(G\), Gene; \(M\), miRNA; \(STE\), seed type effect;
\(RE\), Region Effect; \(E\), Energy; \(STE^\prime\), normalized values
of seed type efficiency coefficient; \(RE^\prime\), normalized values of
region efficiency coefficient; \(E^\prime\), normalized values of energy
coefficient.}\label{fig2}
}
\end{figure}

\iffalse

\%\% (\#TODO the equation is not considering binding factors, right?. We
should explain the calculation made in simulate function here, in simple
sentences. the function looks like this: \%\%
input\_graph\%\textgreater{}\% \%\%
tidygraph::activate(edges)\%\textgreater{}\% \%\%
group\_by(to)\%\textgreater{}\% \%\% mutate(mirna\_count\_per\_comp =
mirna\_count\_current\emph{comp\_count\_current}afff\_factor/sum(comp\_count\_current\emph{afff\_factor),
mirna\_count\_per\_comp = ifelse(is.na(mirna\_count\_per\_comp), 0,
mirna\_count\_per\_comp))\%\textgreater{}\% \%\%
ungroup()\%\textgreater{}\% \%\% mutate(effect\_pre = effect\_current,
effect\_current =
mirna\_count\_per\_comp}degg\_factor)\%\textgreater{}\% \%\%
group\_by(from)\%\textgreater{}\% \%\% mutate(comp\_count\_pre =
ifelse(comp\_count\_current\textless{} 0, 1, comp\_count\_current),
comp\_count\_current = comp\_count\_pre -
(sum(effect\_current)-sum(effect\_pre)), comp\_count\_current =
ifelse(comp\_count\_current\textless{} 0, 1,
comp\_count\_current))\%\textgreater{}\% \%\%
ungroup()\%\textgreater{}\% \%\% mutate(comp\_count\_list =
pmap(list(comp\_count\_list, comp\_count\_current), c), effect\_list =
pmap(list(effect\_list, effect\_current), c), mirna\_count\_list =
pmap(list(mirna\_count\_list, mirna\_count\_current), c)) \%\% ) \fi

I shown in figure and explained at following. I didn't fully understand
whether it is enough or not.

After miRNA binds to its target, but might not repress to bound target.
The functionality of bound miRNA on target depends on efficiency factors
like region that is binding sequence of miRNA on its target. Exact
repression efficiency of miRNA is calculated according to equation
\eqref{eq:4} (\autoref{fig2}d); \(RE^\prime_{gi}\), normalized values of
region efficiency coefficient between miRNA and gene. The cooperative
repression activity of miRNAs to their common targets is figured out as
shown in \autoref{fig2}e.

@@

We have applied our model in a real dataset for which experimental
measurements of various factors were available. Expression levels of
miRNA and genes in tumor and normal tissue of single patient are
retrieved from The Cancer Genome Atlas
\href{https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga}{TCGA}(\#TODO).
High-throughput experimental datasets which are provided miRNA:gene
target pairs with interaction factors (Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}; Moore et al.
\protect\hyperlink{ref-moore_mirnatarget_2015}{2015}). We have combined
miRNA and gene expression datasets via miRNA:target gene dataset
retrieved from \ldots{} (\#FIX name of database and its citation)
answer-\textgreater{}suplementary datasets of (Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}; Moore et al.
\protect\hyperlink{ref-moore_mirnatarget_2015}{2015}). Detailed
description of network construction and its code is available in
Supplementary data (\#TODO burada link nasıl olacak, word dosyası veya
PDF dosyası ismi mi vermemiz mi gerekiyor, yoksa sadece Sup Data
denilmesi yeterli mi? ilgili dosya: TCGA\_E9-A1N5\_article.Rmd). ABCC1
gene, over-expressed in tumor tissue, was selected as trigger for
simulation of integrated dataset. After simulation of network, we have
compared simulation results and tumor tissue expression levels.

@@

\hypertarget{networks-with-single-factor}{%
\subsubsection{Networks with single
factor}\label{networks-with-single-factor}}

We have developed a network-based approach to assess effects of
expression level changes in competitive ceRNA regulation. The basic mode
of miRNA repression activity has been based on miRNA and target
abundance in various researches (Arvey et al.
\protect\hyperlink{ref-arvey_target_2010}{2010}; Denzler et al.
\protect\hyperlink{ref-denzler_assessing_2014}{2014}). Our approach can
effortlessly calculate effects of expression changes when abundance
levels of miRNAs and targets is only available factor. In sample network
given in Figure (\autoref{fig1}), after an increase in expression level
of a gene (G2), expression values of other genes also changed due to
redistribution of miRNA among its targets. Previous studies have shown
that if a gene abundance increases in ceRNA system, expression levels of
genes targeted by shared miRNA are also affected (Lai, Wolkenhauer, and
Vera \protect\hyperlink{ref-lai_understanding_2016}{2016}; Salmena et
al. \protect\hyperlink{ref-salmena_cerna_2011}{2011}; Tay, Rinn, and
Pandolfi \protect\hyperlink{ref-tay_multilayered_2014}{2014}). It was
observed that expression levels of primary neighborhoods which do not
interact with another miRNA of the trigger gene change in relation to
their expression. However, expression level change of common target (G4)
in network has been observed different from G1 that has the same
expression value with G4 target at initial conditions because more than
one miRNA repress this target (\autoref{fig1}d) (\#FIX last sentence
needs some clarification, after ``but cooperative\ldots{}'').

Genes targeted by multiple miRNAs act as a trigger for adjacent local
neighborhood of targeting miRNAs, causing changes in expression levels
of genes outside the local neighborhood of original trigger gene.
Therefore, primary expression level change in gene (G2) causes changes
in other group of genes (G5 and G6) even though original trigger gene
(G2) and genes in other group are not targeted by common miRNA. In
addition, as shown by ceRNA hypothesis model of Ala et al., after the
increase of gene expression level of G2, the miRNA that is found in the
same group (M1) tended to be less repressive on its remaining targets
(G1, G3 and G4). It's important to note that the changes in gene
expression levels will have more pronounced effect if miRNA:target ratio
is high, i.e., more miRNA available per target, which was reported in
previous findings (Arvey et al.
\protect\hyperlink{ref-arvey_target_2010}{2010}; Bosson, Zamudio, and
Sharp \protect\hyperlink{ref-bosson_endogenous_2014}{2014}; Denzler et
al. \protect\hyperlink{ref-denzler_assessing_2014}{2014}). \href{}{}

@@

\hypertarget{competing-endogenous-rnatarget-networks-based-on-interaction-factors}{%
\subsubsection{Competing Endogenous RNA:Target networks based on
interaction
factors}\label{competing-endogenous-rnatarget-networks-based-on-interaction-factors}}

\href{}{\#FIX}: \textless{}\textgreater{} ``title should be more
informative''-\textgreater{} answer.

@@

\leavevmode\hypertarget{ref-atalay2006multidrug}{}%
Atalay, Can, Ismet Deliloglu Gurhan, Cigdem Irkkan, and Ufuk Gunduz.
2006. ``Multidrug Resistance in Locally Advanced Breast Cancer.''
\emph{Tumor Biology} 27 (6): 309--18.

\leavevmode\hypertarget{ref-atalay2008role}{}%
Atalay, C, A Demirkazik, and U Gunduz. 2008. ``Role of Abcb1 and Abcc1
Gene Induction on Survival in Locally Advanced Breast Cancer.''
\emph{Journal of Chemotherapy} 20 (6): 734--39.

@@

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Network based multifactorial modelling microRNA-target interaction.},
            pdfauthor={Selcen Ari; Alper Yilmaz},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Network based multifactorial modelling microRNA-target interaction.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Selcen Ari\footnote{Department of Bioengineering, Yildiz Technical
  University, Istanbul, Turkey} \\ Alper Yilmaz\footnote{Department of Bioengineering, Yildiz Technical
  University, Istanbul, Turkey}}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{15 Eylül 2019}

@@

Competing endogenous RNA (ceRNA) regulations and crosstalk between
various types of non-coding RNA in human is remarkable in means of miRNA
regulation. Many studies have pointed out that an alteration in
miRNA:target interaction can result in unexpected changes due to
indirect and complex interactions. In this paper, we defined a new
network-based model that handles miRNA:ceRNA interactions with
expression values. Our model is able to handle miRNA interaction factors
such as seed type, binding energy, if provided. Our approach is able to
reveal that a perturbation in an element of network affects whole
competing elements differently and cooperative efficiencies of miRNAs on
common targets could be calculated. Our findings emphasized importance
of miRNA:target ratios being crucial, as reported by previous studies.
We have showed that the competing elements which have the same or close
expression values may not be affected equally from the perturbation
because of repression functionality depended on interaction factors of
miRNA target pairs. We applied the model to real sample consisting of
breast cancer gene and miRNA expression dataset and experimental
miRNA:target interaction dataset all generated via high throughput
sequencing methods. A gene over-expressed in tumor tissue, namely
\emph{SERPINE2}, is used as perturbing element. We have observed that
change in expression level of single gene in miRNA:target network is
sufficient to perturb regulations in whole network, due to unforeseen
and unpredicted regulation which are only visible when considered in
network context. Therefore, this model helps unveiling the crosstalk
between elements in miRNA:target network where abundance of target and
sponge effect are taken into account. The model is scalable and can be
plugged in with emerging miRNA effectors such as circRNAs. The model is
available as R package
\href{https://github.com/selcenari/ceRNAnetsim}{ceRNAnetsim}.

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key
regulator of gene expression through various post-transcriptional
mechanisms. Although the mechanisms by which miRNA represses are not
fully understood, miRNAs predominantly repress their targets. Repressive
activities of miRNAs vary depending on many factors that are significant
to miRNA:target interactions. These factors include miRNA:target binding
energy, binding location in target sequence, base pairing types between
miRNA and target, abundance of miRNAs and targets (Grimson et al.
\protect\hyperlink{ref-grimson_microrna_2007}{2007}). Binding energies
of miRNA:target complexes vary based on nucleotide context and determine
folding stability of complex (Cao and Chen
\protect\hyperlink{ref-cao_predicting_2012}{2012}). It has been
demonstrated that the binding energy between miRNA and target indicates
stability or affinity of complex (Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}) and does not directly
determine repressive activity of miRNA (Cao and Chen
\protect\hyperlink{ref-cao_predicting_2012}{2012}). Early studies have
argued that 2-8 nt sequence, seed, located in miRNA 5' end bind to
specific sequence located in 3' UTR of its target (Bartel
\protect\hyperlink{ref-bartel_micrornas_2004}{2004}; Lewis, Burge, and
Bartel \protect\hyperlink{ref-lewis_conserved_2005}{2005}). In recent
studies, it has been shown that miRNAs can interact with targets via
sequences located in regions such as 5' UTR or CDS (Hausser et al.
\protect\hyperlink{ref-hausser_analysis_2013}{2013}; Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}; Moore et al.
\protect\hyperlink{ref-moore_mirnatarget_2015}{2015}). These studies
also showed that binding location either indicate functionality of
miRNA:target interaction or affect abundance of targets. It has been
shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence
that is perfectly complementary with targets, seed region at the 5' end
of miRNAs (Bartel \protect\hyperlink{ref-bartel_micrornas:_2009}{2009};
Grimson et al. \protect\hyperlink{ref-grimson_microrna_2007}{2007}). On
the other hand, some researchers have reported that seed sequence of
miRNA can have mismatches or bulged/wobble nucleotides (Chi, Hannon, and
Darnell \protect\hyperlink{ref-chi2012alternative}{2012}) and may locate
in region other than 5'end of miRNAs (Hafner et al.
\protect\hyperlink{ref-hafner_transcriptome-wide_2010}{2010}; Helwak et
al. \protect\hyperlink{ref-helwak_mapping_2013}{2013}). On top of all
these factors, abundance of miRNAs and targets and miRNA:target ratio in
cells predominantly affect efficiency of miRNA:target interaction (Arvey
et al. \protect\hyperlink{ref-arvey_target_2010}{2010}; Bosson, Zamudio,
and Sharp \protect\hyperlink{ref-bosson_endogenous_2014}{2014}; Denzler
et al. \protect\hyperlink{ref-denzler_assessing_2014}{2014}).

@@

Initially, the network is assumed in steady-state (\autoref{fig1}a)
condition and needs least one trigger for initiating calculations. The
trigger can be a change in expression level of one or more genes
(\autoref{fig1}b). After a trigger, the network undergoes iterative
cycle of calculations at each of which distribution of miRNA in local
neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA
distribution, expression level of each node (i.e.~ceRNA) is updated.
This update results in change of expression value of common target (G4)
in system. In this case, the common target acts as trigger for the other
group (the targets interacted with M2 miRNA) in the network. Due to
common targeted elements, the change in one neighborhood spreads to
other neighborhoods (\autoref{fig1}d), consequently have potential to
affect whole network due to ``ripple effect''.

\begin{figure}
\hypertarget{fig1}{%
\centering
\includegraphics{Fig1.jpg}
\caption{Schematic presentation of mechanism of network based model. a)
In steady state, miRNAs (M: triangles) repress targets (G: circles)
according to proportion of target expression level. b) Two fold increase
in transcript level of Gene2 (G2) acts as a trigger (shown in red). c)
Distribution of miRNA1 (M1) changes. d) The change at expression of
common target affects changes of proportional distribution of miRNA2
(M2). Expression values are rounded to integers for simplicity. Gray
values on edges indicate initial distribution, red values indicate
increase and blue values indicate decrease in distribution. Shades of
circles indicate different levels of increase.}\label{fig1}
}
\end{figure}

@@

The repression efficiency of a miRNA on the individual target
(\(Eff_{gi}\)) is calculated according to equation \eqref{eq:1}; where
miRNA expression (\(C_m\)) in local neighborhood is distributed among
targets using individual gene expression levels (\(C_{gi}\)). For the
genes targeted by multiple miRNAs, cooperative activity of miRNAs on a
target gene, \(R\), is calculated by summing repression activity of each
miRNA (Equation \eqref{eq:2}).

@@

Interactions between miRNAs and their targets can be affected from
various factors. So, our model integrates multiple factors when
calculating overall miRNA activity. We classified factors into two
categories. Binding factors determine interaction between miRNA and
target and they alter amount of miRNA sequestered to target. Efficiency
factors dictate degradation efficieny of miRNA on its target and define
amount of target to be degraded. In other words, binding factors exert
their influence before or during binding, efficiency factors exert their
influence after binding. In the literature, binding free energy (Cao and
Chen \protect\hyperlink{ref-cao_predicting_2012}{2012} ; Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}) and seed type (Werfel
et al. \protect\hyperlink{ref-werfel_preferential_2017}{2017}) in
miRNA:target interactions are described as factors effecting binding
affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to
drastically affect miRNA degradation efficiency (Hausser et al.
\protect\hyperlink{ref-hausser_analysis_2013}{2013}; Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}). Since binding or
degradation efficiency from literature have various and broad ranges,
both binding and efficiency factors are normalized to their maximum
values and scaled to {[}0,1{]} interval. The normalized values of
factors are used to determine binding activity and miRNA efficiency on
targets (\autoref{fig2}). Binding affinities (activity, \(Eff\)) of
miRNAs on each individual gene are calculated as shown in equation
\eqref{eq:3}; \(C_m\), miRNA expression in the group; \(C_{gi}\),
individual gene expression; gi, individual gene (\autoref{fig2}c).
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

\begin{figure}
\hypertarget{fig2}{%
\centering
\includegraphics{fig2.jpg}
\caption{Calculations to determine of miRNA binding and repression
efficiency. \(G\), Gene; \(M\), miRNA; \(STE\), seed type effect;
\(RE\), Region Effect; \(E\), Energy; \(STE^\prime\), normalized values
of seed type efficiency coefficient; \(RE^\prime\), normalized values of
region efficiency coefficient; \(E^\prime\), normalized values of energy
coefficient. Numbering on edges match the pair order in
\autoref{tab:one}}\label{fig2}
}
\end{figure}

Not all miRNA:target binding events result in degradation of target. The
degradation of target by bound miRNA depends on efficiency factors such
as binding region. Exact repression efficiency of miRNA is calculated
according to equation \eqref{eq:4} (\autoref{fig2}d);
\(RE^\prime_{gi}\), normalized values of region efficiency coefficient
between miRNA and gene. The cooperative repression activity of miRNAs to
their common targets is figured out as shown in \autoref{fig2}e.

@@

We have applied our model in a real dataset for which experimental
measurements of various factors were available. Expression levels of
miRNA and genes in tumor and normal tissue of single patient are
retrieved from TCGA Research Network: \url{https://www.cancer.gov/tcga}.
High-throughput experimental datasets which are provided miRNA:gene
target pairs with interaction factors (Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}; Moore et al.
\protect\hyperlink{ref-moore_mirnatarget_2015}{2015}). We have combined
miRNA and gene expression datasets via miRNA:target gene dataset
retrieved from published datasets (Helwak et al.
\protect\hyperlink{ref-helwak_mapping_2013}{2013}; Moore et al.
\protect\hyperlink{ref-moore_mirnatarget_2015}{2015}). Detailed
description of network construction and its code is available in
Supplementary Materials and Methods. We have determined the most
effective nodes in network and fold changes of genes as compared tumor
and normal tissue samples of the same patient.As a result of comparison
of these two datasets, we selected a gene , SERPINE2, as trigger for
simulation of model (see Supplementary Materials and Methods).

@@

\hypertarget{network-with-mirnacerna-expression-level}{%
\subsubsection{Network with miRNA:ceRNA expression
level}\label{network-with-mirnacerna-expression-level}}

We have developed a network-based approach to assess effects of
expression level changes in competitive ceRNA regulation. The basic mode
of miRNA repression activity has been based on miRNA and target
abundance in various researches (Arvey et al.
\protect\hyperlink{ref-arvey_target_2010}{2010}; Denzler et al.
\protect\hyperlink{ref-denzler_assessing_2014}{2014}). Our approach can
effortlessly calculate effects of expression changes when abundance
levels of miRNAs and targets is only available factor. In sample network
given in Figure (\autoref{fig1}), after an increase in expression level
of a gene (G2), expression values of other genes also changed due to
redistribution of miRNA among its targets. Previous studies have shown
that if a gene abundance increases in ceRNA system, expression levels of
genes targeted by shared miRNA are also affected (Lai, Wolkenhauer, and
Vera \protect\hyperlink{ref-lai_understanding_2016}{2016}; Salmena et
al. \protect\hyperlink{ref-salmena_cerna_2011}{2011}; Tay, Rinn, and
Pandolfi \protect\hyperlink{ref-tay_multilayered_2014}{2014}). It was
observed that expression levels of primary neighborhoods which do not
interact with another miRNA of the trigger gene change in relation to
their expression. However, changes in expression levels of G4 and G1
differ although they have same initial conditions. This is due to fact
that G4 is targeted by two miRNAs, M1 and M2 (\autoref{fig1}d).

Genes targeted by multiple miRNAs act as a trigger for adjacent local
neighborhood of targeting miRNAs, causing changes in expression levels
of genes outside the local neighborhood of original trigger gene.
Therefore, primary expression level change in gene (G2) causes changes
in other group of genes (G5 and G6) even though original trigger gene
(G2) and genes in other group are not targeted by common miRNA. In
addition, as shown by ceRNA hypothesis model of Ala et al., after the
increase of gene expression level of G2, the miRNA that is found in the
same group (M1) tended to be less repressive on its remaining targets
(G1, G3 and G4). It's important to note that the changes in gene
expression levels will have more pronounced effect if miRNA:target ratio
is high, i.e., more miRNA available per target, which was reported in
previous findings (Arvey et al.
\protect\hyperlink{ref-arvey_target_2010}{2010}; Bosson, Zamudio, and
Sharp \protect\hyperlink{ref-bosson_endogenous_2014}{2014}; Denzler et
al. \protect\hyperlink{ref-denzler_assessing_2014}{2014}).

@@

\hypertarget{cernatarget-networks-based-on-interaction-factors}{%
\subsubsection{ceRNA:Target networks based on interaction
factors}\label{cernatarget-networks-based-on-interaction-factors}}

@@

@@

\leavevmode\hypertarget{ref-candia2006protease}{}%
Candia, Britny J, William C Hines, Christopher M Heaphy, Jeffrey K
Griffith, and Robert A Orlando. 2006. ``Protease Nexin-1 Expression Is
Altered in Human Breast Cancer.'' \emph{Cancer Cell International} 6
(1): 16.
